Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-26 paper.tex
Move the "Genetic variation" section into tree sequences.

\subsection*{Tree sequences}
\label{sec-ts}
One of the key reasons for \msprime's significiant performance advantage
over other simulators~\citep{kelleher2016efficient}
is its usage of the ``succinct tree sequence''
(usually, tree sequence) data stucture
The basic concepts were introduced by \cite{kelleher2016efficient}
and originally developed as part of \msprime. We have subsequently
extracted the tree sequence code from \msprime\ into
\tskit~\citep{tskit2021tskit}, a standalone Python and C library.
While a full discussion of tree sequences and the capabilities
of \tskit\ is beyond the scope of this article, we summarise
the aspects that are important for simulation here.

@@

The succinct tree sequence is a data structure for concisely encoding
genetic ancestry and variation. Given a set of sample chromosomes, the topology
of their genetic ancestry is described by a set of \emph{nodes}
and \emph{edges} (Fig.~\ref{fig-ts-example}).
A node in tree sequence terminology corresponds
to a single chromosome strand, and is associated with its birth time,
population and other metadata. A node may optionally be
associated with an \emph{individual}, so that, for example, a diploid
individual will be associated with two separate nodes (we omit
the individual table in Fig.~\ref{fig-ts-example} as it is optional).
Nodes define haplotypes that exist at a specific time within a
some individual, and edges define genetic inheritence
relationships between these nodes. An edge consists of a
parent node, a child node and the left and right coordinates
of the contiguous genomic interval over which
the child inherited genetic material from the parent.
Critically, edges can span multiple marginal trees: for
example, in Fig.~\ref{fig-ts-example} the branch from node
0 to 4 is present in both marginal trees, and represented
by a single edge (the first row in the example edge table).
This simple device, of explicitly associating tree nodes
with specific ancestral chromosomes and recording the
genomic intervals over which parent/child relationships exist,
is a generalisation from the original concept  of
``coalescence records''~\citep{kelleher2016efficient}, and
is the key to the efficiency of tree
sequences~\citep{
kelleher2018efficient,kelleher2019inferring,ralph2020efficiently}.
The term Ancestral Recombination Graph (ARG) is often understood
to mean the complete record of the genetic
ancestry of a set of sampled
genomes~\citep[e.g.][]{mathieson2020ancestry},
and in this sense a tree
sequence \emph{is} an ARG. More technically, an ARG is a particular
graphical representation of genetic ancestry based on the
flow of ancestral material through recombination and common
ancestor events~\citep{griffiths1991two,griffiths1997ancestral}.
In this sense, tree sequences can be used to \emph{encode}
ARGs without loss of information (although we do not store
the additional ARG information by default
in \msprime\ because the minimal tree sequence encoding is
much more concise; see the \nameref{sec-arg} section for more details).
It should be noted that, despite efforts to
standardise~\citep{cardona2008extended,mcgill2013graphml}, there is
no shared format for interchanging ARG data, and no supporting software
libraries.

\subsection*{Genetic variation}
\label{sec-genetic-variation}

The output of population genetics simulators, quite naturally,
tends to be some representation of the resulting nucleotide sequences
(or some derived statistics of those sequences),
although the underlying genealogies are also of interest.
% Mutations are not treated as first-class data --- the sequences
% were really all people were after, classically.
However, joint information about the mutations that have been simulated
-- e.g., where they have occurred on the trees --
is usually not retained or made available for subsequent analysis.

The approach taken in \msprime\ is quite different.
As illustrated in Fig.~\ref{fig-ts-example}, mutations are an
integral part of \tskit's tree sequence data model, and genetic
variation is encoded by recording sites at which mutations
have occurred, and where each mutation at those sites
has occurred on the marginal tree;
the sequences themselves are never stored.
It may at first seem inconvenient to have only this indirect representation
of the genome sequences, but it is extremely powerful.
Firstly, the storage space required for simulations is dramatically
reduced. For a simulation of $n$ samples with $m$ variant sites,
we would require $O(nm)$ space to store the sequence data as a variant
matrix. However, if this simulation was of a recombining genome
with $t$ trees, then the \tskit\ tree sequence encoding
requires $O(n + t + m)$ space~\citep{kelleher2016efficient},
assuming we have $O(1)$ mutations at each site. For
large sample sizes, this difference is profound, making it
conceivable, for example, to store the genetic ancestry
and variation data for the entire human population on a
laptop~\citep{kelleher2019inferring}. Besides the huge
difference in storage efficiency, it is also often far more efficient
to compute statistics of the sequence data from the
trees and mutations than it is to work with the sequences themselves.
For example, computing Tajima's $D$ from simulated data using \tskit\
is several orders of magnitude faster than efficient variant matrix
libraries for large sample sizes~\citep{ralph2020efficiently}.

The advantages of storing genetic variation data using the underlying trees
have recently become clear in the context of the vast genomic
datasets produced during the SARS-CoV-2 pandemic.
\cite{turakhia2021ultrafast} propose the
Mutation Annotated Tree (MAT)
format (consisting of a Newick tree
and associated mutations in a binary format) and
the \texttt{matUtils} program as an efficient way to store
and process large viral datasets~\citep{mcbroome2021daily},
achieving excellent compression and processing performance.
Similarly, \phastsim~\citep{demaio2021phastsim}
was developed to simulate sequence evolution on such large SARS-CoV-2
phylogenies, and also outputs a Newick tree annotated with mutations (not
in MAT format) to avoid the bottleneck of generating
and storing the simulated sequences.
While these methods illustrate the advantages of the general
approach of storing ancestry and mutations rather than sequences,
they do not generalise beyond their immediate settings, and
no software library support is available.
\msprime, on the other hand, provides a fast, powerful, and well-tested
generic interface for generating and storing mutations on trees.

@@

% Explain the background: what have people done classically?
Because coalescent simulations are usually concerned with
neutral evolution (see the \nameref{sec-selection} section, however)
the problem of generating synthetic genetic variation can be decomposed into
two independent steps:
simulating genetic ancestry (the trees), and then simulating
variation by superimposing mutation processes on those trees
(see figure~\ref{fig-mutated-trees}).
A number of programs exist to place mutations on trees:
for instance,
the classical \texttt{Seq-Gen} program~\citep{rambaut1997seq}
allows a range of different models of sequence evolution to be imposed on a given
tree in Newick format, and there are various
extensions to the basic nucleotide
models~\citep[e.g.][]{cartwright2005dna,fletcher2009indelible}.
Population genetic simulators have tended to
include their own implementations of mutation simulation, with
most just supporting the infinite sites
model~\citep[e.g.][]{hudson2002generating}
but with several supporting a range of different models
of sequence
evolution~\citep[e.g.][]{mailund2005coasim,excoffier2011fastsimcoal,
virgoulay2021gspace}.

@@

\begin{figure}
\begin{center}
    \includegraphics{figures/ancestry-perf}
\end{center}
\caption{\label{fig-ancestry-perf}
    Approximate run time for \msprime\ simulations,
    for \textbf{(left)} small, and \textbf{(right)} larger simulations.
    Each point is the runtime of one simulation,
    for various values of effective population size ($N$),
    genome length ($L$), and number of samples.
    Runtime scales quadratically with the product of $N$ and $L$;
    since recombination rate in these simulations was $10^{-8}$,
    the scaled recombination rate here is $\rho = NL/10^8$,
    shown on the horizontal axis.
    The black line is a quadratic fit separately in each panel.
    \plrcomment{TODO: run out more simulations to get more triangles}
}
\end{figure}

@@

Figure~\ref{fig-ancestry-perf} gives a rough idea of the runtime of
a given simulation of a constant size population,
and verifies that runtime scales quadratically with $\rho$,
and depends much less strongly on the number of samples.
Taking a typical chromosome to be 1 Morgan in length,
these plot show, roughly, that simulating
samples from a population of thousands of individuals takes seconds,
while samples from a population of tens of thousands takes minutes.

However, what about a simulation with \emph{changing} size?
To understand how runtime depends on demography
it helps to consider \emph{why} runtime is quadratic in $\rho$.
At any point in time, \msprime\ must keep track of some number of lineages,
each of which contains some number of chunks of genetic material.
Common ancestor events reduce the number of lineages,
and recombination events increase their number.
However, with long genomes
only a small fraction of the common ancestor events
will involve overlapping segments of ancesty,
and lead to coalescence in the marginal trees.
Such disjoint segments are often far apart (on average, about distance $L/2$),
and so recombine apart again immediately;
it is these large numbers of rapid yet inconsequential events that lead to the quadratic runtime.
\plrcomment{(this is discussed also in the ARG section; check for overlap)}
The maximum number of lineages occurs when
the increase and decrease in numbers of lineages due to
common ancestor and recombination events balance out;
to get an idea of runtime we can estimate when this occurs.
Suppose that the maximal number of lineages is $M$;
at this time the rate of common ancestor events is $M (M-1) / (4 N_e)$
and the total rate of recombination is $M \ell$,
where $\ell$ is the mean length of genome carried by each lineage
(including ``trapped'' non-ancestral material).
It is known that the maximum number of lineages is reached very quickly \citep{nelson2020accounting},
and at this maximum, recombination and coalescence rates are equal,
so many (roughly half) of these lineages have at least two distant segments of ancestral material
separated by a long segment of genome, implying that $\ell \approx L / 4$.
This implies that $M$ is roughly $L N_e$.
Now, the number of linages decreases gradually, on the coalescent time scale -- so, over roughly $2 N_e$ generations.
Since the total rate of events when there are the maximum number of lineages
happens at rate roughly $L^2 N_e / 4$,
then the total number of events is proportional to $(L N_e)^2$ -- i.e., proportional to $\rho^2$,
as desired.
What does this tell us about time-varying population sizes?
The argument above implies that the work is spread out relatively evenly on the coalescent time scale:
in other words,
more events happen in the time period from 0 to $N_e/2$ generations
than from $N_e/2$ to $N_e$ generations,
but the runtime is not dominated but the more recent time period.

Putting this together:
suppose that population size today is $N_1$,
while longer ago than $T$ generations ago it was $N_2$:
does the runtime depend more on $4 N_1 L$ or $4 N_2 L$?
The answer depends on how $T$ commpares to $N_1$: if $T/N_1$ is large,
then runtime will be similar to a population of size $N_1$;
while if $T/N_1$ is small, it will be similar to a population of size $N_2$.
For instance, in many agricultural species $N_1 \propto 100$, while $N_2 \propto 10^5$,
and the runtime will depend critically on $T$ -- in other words,
simulation will be quick in a species with a strong domestication bottleneck,
and slow otherwise.
In practice, we recommend that users estimate runtime by
measuring runtime in a series of simulations with short genome lengths,
possibly varying other parameters to check for interactions,
and then predict runtime by fitting a quadratic curve to genome length.

@@

@@

\label{app-likelihoods}
\section*{Likelihood calculations}

>>>>

\subsection*{Tree sequences}
\label{sec-ts}

@@

One of the key reasons for \msprime's substantial performance advantage
over other simulators~\citep{kelleher2016efficient}
is its usage of the ``succinct tree sequence''
 data stucture to represent simulation results.
The succinct tree sequence (usually abbreviated to tree sequence)
is a data structure for concisely encoding
genetic ancestry and sequence variation.
The basic concepts were introduced by \cite{kelleher2016efficient}
and originally developed as part of \msprime. We subsequently
extracted the tree sequence code that was independent of simulations
from \msprime\ into \tskit,
a standalone Python and C library~\citep{tskit2021tskit}.
While a full discussion of tree sequences and the capabilities
of \tskit\ is beyond the scope of this article, we summarise
some aspects that are important for simulation here.

% Encoding ancestry
Given a set of sample chromosomes, the topology
of their genetic ancestry is described by a set of \emph{nodes}
and \emph{edges} (Fig.~\ref{fig-ts-example}).
A node in tree sequence terminology corresponds
to a single chromosome strand, and is associated with its birth time,
and other optional metadata. A node may optionally be
associated with an \emph{individual}, so that, for example, a diploid
individual will be associated with two separate nodes (we omit
the individual table in Fig.~\ref{fig-ts-example} as it is optional).
Nodes define haplotypes that exist at a specific time within
some individual, and edges define genetic inheritence
relationships between these nodes. An \emph{edge} consists of a
parent node, a child node and the left and right coordinates
of the contiguous genomic interval over which
the child inherited genetic material from the parent.
Critically, edges can span multiple marginal trees: for
example, in Fig.~\ref{fig-ts-example} the branch from node
0 to 4 is present in both marginal trees, and represented
by a single edge (the first row in the edge table).
This simple device, of explicitly associating tree nodes
with specific ancestral chromosomes and recording the
genomic intervals over which parent/child relationships exist,
generalises the original
``coalescence records'' concept~\citep{kelleher2016efficient},
and is the key to the efficiency of tree
sequences~\citep{
kelleher2018efficient,kelleher2019inferring,ralph2020efficiently}.

% Quick digression: relationship to ARGs
The term Ancestral Recombination Graph (ARG) is often understood
to mean the complete record of the genetic
ancestry of a set of sampled
genomes~\citep[e.g.][]{mathieson2020ancestry},
and in this sense a succint tree
sequence \emph{is} an ARG. More technically, an ARG is a particular
graphical representation of genetic ancestry based on the
flow of ancestral material through recombination and common
ancestor events~\citep{griffiths1991two,griffiths1997ancestral}.
In this sense, tree sequences can be used to \emph{encode}
ARGs without loss of information (although
the default minimal tree sequence encoding is
much more concise; see the \nameref{sec-arg} section).
It should be noted that, despite efforts to
standardise~\citep{cardona2008extended,mcgill2013graphml}, there is
no shared format for interchanging ARG data, and no supporting software
libraries.

% OK, back to tree sequences: what about sequence variation?
The final output of most population simulations is some representation
of the sequence variation among the specified samples. For coalescent
simulations, we usually have three steps (1) simulate
the genetic ancestry, and optionally output the resulting marginal trees;
(2) simulate sequence evolution conditioned on this ancestry by generating
mutations (see the \nameref{sec-mutations} section); and (3) output the
resulting sequences by percolating the effects of the mutations through
the trees. Information about the mutations
themselves---e.g., where they have occurred on the
trees---is usually not retained or made available for subsequent analysis.
In \msprime, however, we skip step (3), instead using \tskit's
data model of ancestry and mutations
to represent the simulated sequences.
As illustrated in Fig.~\ref{fig-ts-example}, mutations are a
fully integrated part of \tskit's tree sequence data model, and genetic
variation is encoded by recording sites at which mutations
have occurred, and where each mutation at those sites
has occurred on the marginal tree.
Crucially, the sequences themselves are never stored, or indeed directly
represented in memory (unless explicitly decoded using \tskit\ APIs).
It may at first seem inconvenient to have only this indirect representation
of the genome sequences, but it is extremely powerful.
Firstly, the storage space required for simulations is dramatically
reduced. For a simulation of $n$ samples with $m$ variant sites,
we would require $O(nm)$ space to store the sequence data as a variant
matrix. However, if this simulation was of a recombining genome
with $t$ trees, then the \tskit\ tree sequence encoding
requires $O(n + t + m)$ space,
assuming we have $O(1)$ mutations at each site~\citep{kelleher2016efficient}.
For large sample sizes, this difference is profound, making it
conceivable, for example, to store the genetic ancestry
and variation data for the entire human population on a
laptop~\citep{kelleher2019inferring}. Besides the huge
difference in storage efficiency, it is also often far more efficient
to compute statistics of the sequence data from the
trees and mutations than it is to work with the sequences themselves.
For example, computing Tajima's $D$ from simulated data using \tskit\
is several orders of magnitude faster than efficient variant matrix
libraries for large sample sizes~\citep{ralph2020efficiently}.

The advantages of storing genetic variation data using the underlying trees
have recently become clear in the context of the vast genomic
datasets produced during the SARS-CoV-2 pandemic.
\cite{turakhia2021ultrafast} propose the
Mutation Annotated Tree (MAT)
format (consisting of a Newick tree
and associated mutations in a binary format) and
the \texttt{matUtils} program as an efficient way to store
and process large viral datasets~\citep{mcbroome2021daily},
achieving excellent compression and processing performance.
Similarly, \phastsim~\citep{demaio2021phastsim}
was developed to simulate sequence evolution on such large SARS-CoV-2
phylogenies, and also outputs a Newick tree annotated with mutations (not
in MAT format) to avoid the bottleneck of generating
and storing the simulated sequences.
While these methods illustrate the advantages of the general
approach of storing ancestry and mutations rather than sequences,
they do not generalise beyond their immediate settings, and
no software library support is available.
% \msprime, on the other hand, provides a fast, powerful, and well-tested
% generic interface for generating and storing mutations on trees.

@@

% Explain the background: what have people done classically?
Because coalescent simulations are usually concerned with
neutral evolution (see the \nameref{sec-selection} section, however)
the problem of generating synthetic genetic variation can be decomposed into
two independent steps:
simulating genetic ancestry (the trees), and then simulating
variation by superimposing mutation processes on those trees
(see Fig.~\ref{fig-mutated-trees}).
A number of programs exist to place mutations on trees:
for instance,
the classical \texttt{Seq-Gen} program~\citep{rambaut1997seq}
allows a range of different models of sequence evolution to be imposed on a given
tree in Newick format, and there are various
extensions to the basic
models~\citep[e.g.][]{cartwright2005dna,fletcher2009indelible}.
Population genetic simulators have tended to
include their own implementations of mutation simulation, with
most supporting the infinite sites
model~\citep[e.g.][]{hudson2002generating}
but with several supporting a range of different models
of sequence
evolution~\citep[e.g.][]{mailund2005coasim,excoffier2011fastsimcoal,
virgoulay2021gspace}.

@@

\begin{figure}
\begin{center}
    \includegraphics{figures/ancestry-perf}
\end{center}
\caption{\label{fig-ancestry-perf}
    Approximate run time for \msprime\ simulations,
    for \textbf{(left)} small, and \textbf{(right)} larger simulations.
    Each point is the runtime of one simulation,
    for various values of effective population size ($N$),
    genome length ($L$), and number of samples.
    Runtime scales quadratically with the product of $N$ and $L$;
    since recombination rate in these simulations was $10^{-8}$,
    the scaled recombination rate here is $\rho = NL/10^8$,
    shown on the horizontal axis.
    The black line is a quadratic fit separately in each panel.
}
\end{figure}

@@

Figure~\ref{fig-ancestry-perf} gives a rough idea of the runtime of
a given simulation of a constant size population,
and verifies that runtime scales quadratically with $\rho$,
(\citet{hein2004gene} gives an equation
including the effects of sample size --
see Appendix~\ref{app-theoretical-runtime}.)
and depends much less strongly on the number of samples.
Taking a typical chromosome to be 1 Morgan in length,
these plot show, roughly, that simulating
samples from a population of thousands of individuals takes seconds,
while samples from a population of tens of thousands takes minutes.

However, what about a simulation with \emph{changing} size?
To understand how runtime depends on demography
it helps to consider \emph{why} runtime is quadratic in $\rho$.
At any point in time, \msprime\ must keep track of some number of lineages,
each of which contains some number of chunks of genetic material.
Common ancestor events reduce the number of lineages,
and recombination events increase their number.
However, with long genomes
only a small fraction of the common ancestor events
will involve overlapping segments of ancestry,
and lead to coalescence in the marginal trees.
Such disjoint segments are often far apart (on average, about distance $L/2$),
and so recombine apart again immediately;
it is these large numbers of rapid yet inconsequential events that lead to the quadratic runtime.
\plrcomment{(this is discussed also in the ARG section; check for overlap)}
The maximum number of lineages occurs when
the increase and decrease in numbers of lineages due to
common ancestor and recombination events balance out;
to get an idea of runtime we can estimate when this occurs.
Suppose that the maximal number of lineages is $M$;
at this time the rate of common ancestor events is $M (M-1) / (4 N_e)$
and the total rate of recombination is $M \ell$,
where $\ell$ is the mean length of genome carried by each lineage
(including ``trapped'' non-ancestral material).
It is known that the maximum number of lineages is reached very quickly \citep{nelson2020accounting},
and at this maximum, recombination and coalescence rates are equal,
so many (roughly half) of these lineages have at least two distant segments of ancestral material
separated by a long segment of genome, implying that $\ell \approx L / 4$.
This implies that $M$ is roughly $L N_e$.
Now, the number of linages decreases gradually, on the coalescent time scale -- so, over roughly $2 N_e$ generations.
Since the total rate of events when there are the maximum number of lineages
happens at rate roughly $L^2 N_e / 4$,
then the total number of events is proportional to $(L N_e)^2$ -- i.e., proportional to $\rho^2$,
as desired.
What does this tell us about time-varying population sizes?
The argument above implies that the work is spread out relatively evenly on the coalescent time scale:
in other words,
more events happen in the time period from 0 to $N_e/2$ generations
than from $N_e/2$ to $N_e$ generations,
but the runtime is not dominated but the more recent time period.

Putting this together:
suppose that population size today is $N_1$,
while longer ago than $T$ generations ago it was $N_2$:
does the runtime depend more on $4 N_1 L$ or $4 N_2 L$?
The answer depends on how $T$ compares to $N_1$: if $T/N_1$ is large,
then runtime will be similar to a population of size $N_1$;
while if $T/N_1$ is small, it will be similar to a population of size $N_2$.
For instance, in many agricultural species $N_1 \propto 100$, while $N_2 \propto 10^5$,
and the runtime will depend critically on $T$ -- in other words,
simulation will be quick in a species with a strong domestication bottleneck,
and slow otherwise.
In practice, we recommend that users estimate runtime by
measuring runtime in a series of simulations with short genome lengths,
possibly varying other parameters to check for interactions,
and then predict runtime by fitting a quadratic curve to genome length.
There is only a weak dependency on sample size
-- notice that in Figure~\ref{fig-ancestry-perf} increasing sample size 100-fold
only increases runtime by a factor of 2 or so.

@@

\subsection*{Analysis of simulation results}
\label{sec-data-analysis}

\jkcomment{This is probably part of the discussion, but putting here out of
the way for now in its own subsection}

@@

\section*{Theoretical runtime}
    \label{app-theoretical-runtime}

\begin{figure}
\begin{center}
    \includegraphics{figures/ancestry-perf-expected}
\end{center}
\caption{\label{fig-expected-perf}
    Data from Figure~\ref{fig-ancestry-perf}, compared with
    equation~\eqref{eqn-hsw-bound} (``expected'').
    Left panel shows observed times plotted against expected number
    of operations; right panel shows the ratio of the two, plotted against $\rho$.
}
\end{figure}

The book \citet{hein2004gene} (Equation 5.10) gives a theoretical upper bound
on the number of events involving trapped ancestral material in Hudson's algorithm.
For $n$ samples from 
a diploid population of effective size $N_e$,
and a genome of genetic length $r$,
with $\rho = 4 N_e r$ the scaled recombination rate,
and $R^T$ the number of recombination recombination events in trapped material,
this upper bound is
\begin{equation} \label{eqn-hsw-bound}
    \mathbb{E}[R^T]
    \le
    \rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2 .
\end{equation}
Note that the sum on the right is well-approximated by $\log(n)$.
Although \citet{hein2004gene} refer to this bound as ``crude'',
it turns out to predict the scaling of runtime quite well for large genomes.
Figure~\ref{fig-expected-perf} shows the observed runtime
(from the same simulations as Figure~\ref{fig-ancestry-perf})
plotted against equation~\eqref{eqn-hsw-bound},
showing that for values of $\rho$ above $4 \times 10^4$ or so
(i.e., chromosome-length simulations in populations above $10^4$),
the ratio of observed time to the value from \eqref{eqn-hsw-bound} is constant.
This implies the runtime for a large sample
can be predicted from the runtime for a small sample.
For smaller values of $\rho$, however, the prediction
severely underestimates runtime.

\section*{Likelihood calculations}
\label{app-likelihoods}
